The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-06 manuscript/manuscript.Rnw
in bos taurus, the gene is HSD3B not HSD3B1.
fix elephant table label

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elphant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB10} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of the annexin complex were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B1} %
(\Sexpr{gene.expression("bos taurus","HSD3B1")}).

@@

\begin{table}
\tiny
<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
print(xtable(top.elephant.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.elephant.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Loxodonta africana}}
\label{tab:top50_elphant_placenta}
\end{table}

>>>>

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB10} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of the annexin complex were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

\begin{table}
\tiny
<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
print(xtable(top.elephant.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.elephant.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Loxodonta africana}}
\label{tab:top50_elephant_placenta}
\end{table}
